# Reproducible Research: Peer Assessment 2

```{r}
library("ggplot2")
library("reshape2")
```

## Loading the data

The data for this assignment come in the form of a comma-separated-value file 
compressed via the bzip2 algorithm to reduce its size.  
It is sourced from the Coursera site.


```{r}
#Download file and decompress
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
file <- basename(url)
download.file(url, file, method="curl")
fcon <- bzfile(file, open="r")
#Read CSV file
stormData <-read.csv(fcon, stringsAsFactors=FALSE)
close(fcon)
```

## Categorize and process the data
```{r}
#Keep relevant columns
stormData <- stormData[, c("BGN_DATE", "STATE", "EVTYPE", "END_DATE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
#Remove non US-states such as PR and ST in the data
stormData <- stormData[(stormData$STATE %in% c(state.abb, "DC")), ]
#Convert END_DATE and BGN_DATE to dates
stormData$BGN_DATE <- as.Date(stormData$BGN_DATE, format="%m/%d/%Y")
stormData$END_DATE <- as.Date(stormData$END_DATE, format="%m/%d/%Y")
#Lowercase EVTYPE
stormData$EVTYPE <- tolower(stormData$EVTYPE)
#Remove non-pertinent Weather events (no human, property, or crop damage)
stormData <- stormData[!(stormData$FATALITIES == 0 & stormData$INJURIES == 0 & stormData$PROPDMG == 0 & stormData$CROPDMG == 0),]

#Categorize the events from least specific to most
stormData[grepl("turbulence|apache|urban|normal|hyperthermia|record temperatures|other|dam|southeast|beach|drowning|none|severe|\\?|mild|excessive|record", stormData$EVTYPE), "Category"] <- "Other Events"
stormData[grepl("dry|drought", stormData$EVTYPE), "Category"] <- "Very Dry/Drought"
stormData[grepl("marine|rough|southeast|seiche|stream|heavy seas|wave", stormData$EVTYPE), "Category"] <- "Sea/Lake/River Events"
stormData[grepl("summary|metro", stormData$EVTYPE), "Category"] <- "Summary of Date"
stormData[grepl("warm|heat|hot|high|warmth", stormData$EVTYPE), "Category"] <- "Heat Events"
stormData[grepl("heavy mix|snow|sleet|wintry|winter|cold|cool|freezing|blizzard|freeze|low", stormData$EVTYPE), "Category"] <- "Cold/Snow/Winter Weather/Blizzard"
stormData[grepl("rainfall|rain|precip|precipatation|precipitation|wet", stormData$EVTYPE), "Category"] <- "Rain & Percipatation"
stormData[grepl("fog", stormData$EVTYPE), "Category"] <- "Fog"
stormData[grepl("cloud", stormData$EVTYPE), "Category"] <- "Cloud Related Events"
stormData[grepl("dust", stormData$EVTYPE), "Category"] <- "Dust Related Events"
stormData[grepl("floodin|stream fld|flood|rapidly|flash", stormData$EVTYPE), "Category"] <- "Flood"
stormData[grepl("torndao|tornado|funnel|spout|gustnado", stormData$EVTYPE), "Category"] <- "Tornado/Gustnado/Funnel Cloud/Waterspout"
stormData[grepl("avalance|avalanche", stormData$EVTYPE), "Category"] <- "Avalanche"
stormData[grepl("coastalstorm|coastal storm", stormData$EVTYPE), "Category"] <- "Coastal Storm"
stormData[grepl("thunderstorm|tstm|thunderstormw|lightning|lighting|ligntning", stormData$EVTYPE), "Category"] <- "Thunderstorms/Lightning"
stormData[grepl("glaze|frost|ice|icy", stormData$EVTYPE), "Category"] <- "Ice Events"
stormData[grepl("wind|burst", stormData$EVTYPE), "Category"] <- "Wind, Microbursts, and Downburst Events"
stormData[grepl("hail", stormData$EVTYPE), "Category"] <- "Hail Events"
stormData[grepl("hurricane", stormData$EVTYPE), "Category"] <- "Hurricanes"
stormData[grepl("current|tide|surf|surge", stormData$EVTYPE), "Category"] <- "Currents, Surfs, and Tides"
stormData[grepl("fires|wildfire|fire|smoke", stormData$EVTYPE), "Category"] <- "Fires/Smoke"
stormData[grepl("mud|rock|mudslides|landslide|landslump|erosion", stormData$EVTYPE), "Category"] <- "Rock/Mud/Landslides & Erosion"
stormData[grepl("tropical storm|depression", stormData$EVTYPE), "Category"] <- "Tropical Storms/Depressions"
stormData[grepl("tsunami", stormData$EVTYPE), "Category"] <- "Tsunami"
```


##Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Given the graph below, torandic events are clearly most harmful to population health.
```{r}
library(grid)
library(gridExtra)

#Plot All Events by Category
stormHealth <- aggregate(cbind(FATALITIES, INJURIES) ~ Category, sum, data = stormData)
stormHealth <- stormHealth[!(stormHealth$FATALITIES == 0 & stormHealth$INJURIES == 0),]
meltStormHealth <- melt(stormHealth, id="Category", variable.name="Health_Type", value.name="Number_Of_Incidents")
p1 <- qplot(x=Category, y=Number_Of_Incidents, fill=Health_Type, data=meltStormHealth, geom="bar", stat="identity") + coord_flip()

#Break down the "Other Category"
stormOtherHealth <- stormData[(stormData$Category == "Other Events"),]
stormOtherHealth <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, sum, data = stormOtherHealth)
stormOtherHealth <- stormOtherHealth[!(stormOtherHealth$FATALITIES == 0 & stormOtherHealth$INJURIES == 0),]
meltStormOtherHealth <- melt(stormOtherHealth, id="EVTYPE", variable.name="Health_Type", value.name="Number_Of_Incidents")
p2 <- qplot(x=EVTYPE, y=Number_Of_Incidents, fill=Health_Type, data=meltStormOtherHealth, geom="bar", stat="identity") + coord_flip() + xlab("Event Type")

grid.arrange(p1, p2, ncol = 1, main = "Health Events (Left= All Event Types, Right=Fine Grain Detail of Other category)"
```

As shown in the table below, see the highest number of incidents by type
```{r, results='asis'}
library(knitr)
##Create two data frames, fatalities and injuries sorted by the highest counts at the top
fatalities <- meltStormHealth[meltStormHealth$Health_Type == "FATALITIES",]
fatalities <- fatalities[order(fatalities$Number_Of_Incidents, decreasing=TRUE),]
injuries <- meltStormHealth[meltStormHealth$Health_Type == "INJURIES",]
injuries <- injuries[order(injuries$Number_Of_Incidents, decreasing=TRUE),]
t1 = kable(fatalities, format='html', output = FALSE)
t2 = kable(injuries, format='html', output = FALSE)
cat(c('<table><tr valign="top"><td>', t1, '</td><td>', t2, '</td><tr></table>'),
    sep = '')
```
